Research Proposal: Specialization Trends in Economics Research using JEL Codes \\ Text Data in Economics by Aapo Stenhammar \\ University of Bonn, Winter Term 2024/25

### 2025-02-27 documents/paper.tex
Paper writing.

\clearpage

@@

\subsection{Descriptive Statistics}

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{../bld/figures/fig_counts_avg_jel_codes}
    \caption{Discussion Papers \& Avg. JEL Codes per Paper Over Time}
    \label{fig:monthly_counts_and_avg_jel_codes_per_paper}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{../bld/figures/fig_author_trends}
    \caption{Average Number of Authors Over Time}
    \label{fig:author_trends}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{../bld/figures/fig_dp_counts}
    \caption{Number of Authors \& JEL Codes per Discussion Paper}
    \label{fig:dp_counts}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{../bld/figures/fig_top5overall_jel}
    \caption{Top 5 JEL Codes}
    \label{fig:top5overall_jel}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{../bld/figures/fig_top3yearly_jel}
    \caption{Top 3 JEL Codes yearly}
    \label{fig:top3yearly_jel}
\end{figure}

\section{Methodology}

\subsection{Topic Modeling}
To analyze the specialization trends in economics research, topic modeling is used.
Specifically, the Latent Dirichlet Allocation (LDA) algorithm is used to identify topics in the abstracts of the discussion papers \cite{blei2003}.
To prepare the abstract data for the LDA model, the text is lowercased, tokenized, and stemmed.
Numbers, punctuation, and words containing no alphanumeric characters are removed.
Furthermore, stopwords are removed from the text.
The model
The number of topics is set to 750 and 20 passes through the corpus of tokenized and stemmed abstracts are performed during training of the model, similar to \cite{galiani2023a}.

@@

@@

@@

>>>>

@@

\subsection{Descriptive Statistics}
To get an overview of the data, the following descriptive statistics are presented.
Statistics of interest are related to discussion papers, JEL Codes, and authors.
In Figure \ref{fig:dp_counts}, it can be seen that the number of discussion papers published per month has increased over time from less than 20 to over 80 per month.
Interesting to observe is the positive impact of the initial period of the COVID-19 pandemic on the number of discussion papers published.
The average number of JEL Codes per discussion paper on the other hand is rather stable over time, and fluctuates only mildly.

\begin{figure}
    \centering
    \includegraphics{../bld/figures/fig_dp_counts}
    \caption{Number of Discussion Papers \& JEL Codes per Discussion Paper}
    \label{fig:dp_counts}
\end{figure}

Relevant to note is, that the structure of the authors publishing discussion papers has changed over time.
The average number of authors per discussion paper per month has increased by about 0.5 authors per discussion paper.
Furthermore, the average new number of new authors per discussion paper per month is stable around 0.5 authors per discussion paper, while the average number of authors that have published before is increasing by about 1.5 authors per discussion paper per month.
This fact can be related to the establishment of the discussion series as a well-known platform for research in Labor Economics.
These trends are shown in Figure \ref{fig:author_trends}.

\begin{figure}
    \centering
    \includegraphics{../bld/figures/fig_author_trends}
    \caption{Average Number of Authors Over Time}
    \label{fig:author_trends}
\end{figure}

\section{Methodology}
The methodologies presented here are used to achieve two main objectives.
First, the behavior of subfields over time is to be analyzed.
Second, the text data methods, such as topic modeling using Latent Dirichlet Allocation (LDA), are used to validate the JEL Codes as a tool for analyzing specialization trends.

\subsection{Topic Modeling}
To analyze the specialization trends in economics research, topic modeling is used.
Specifically, the LDA algorithm is used to identify topics in the abstracts of the discussion papers \cite{blei2003}.
To prepare the abstract data for the LDA model, the text is lowercased, tokenized, and stemmed.
Numbers, punctuation, and words containing no alphanumeric characters are removed.
Furthermore, stopwords are removed from the text.
The model
The number of topics is set to 750 and 20 passes through the corpus of tokenized and stemmed abstracts are performed during training of the model, similar to \cite{galiani2023a}.

@@

In Figure \ref{fig:top5_increasing_decreasing_topics}, the top 5 topics increasing and decreasing in presence are shown.

\begin{figure}
    \centering
    \includegraphics{../bld/figures/fig_topic_trends}
    \caption{Top 5 Topics Increasing (Left) and Decreasing (Right) in Presence}
    \label{fig:top5_increasing_decreasing_topics}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics{../bld/figures/fig_top5overall_jel}
    \caption{Top 5 JEL Codes}
    \label{fig:top5overall_jel}
\end{figure}

In the Figure \ref{fig:top3yearl_jel}, all JEL Codes are shown that are at least in one year among the top 3 JEL Codes.
If the JEL Code is not among the top 3 JEL Codes in a year, the opacity of the line is reduced.
The JEL Codes 'J24' (Human Capital, Skills, Occupational Choice, Labor Productivity) and 'J31' (Wage Level and Structure, Wage Differentials) are consistently among the top 3 JEL Codes.
While other JEL Codes like 'J16' (Economics of Gender, Non-labor Discrimination) saw a relative frequency increase lately.

\begin{figure}
    \centering
    \includegraphics{../bld/figures/fig_top3yearly_jel}
    \caption{Top 3 JEL Codes yearly}
    \label{fig:top3yearly_jel}
\end{figure}

@@

\section{Appendix}

There are few authors that have published a large number of discussion papers.
Similarly, there are few JEL Codes that are assigned to a large number of discussion papers.
The Figure \ref{fig:counts_dp_jel_codes} shows the distribution of discussion papers by author on the left and by JEL Codes on the right.

\begin{figure}
    \centering
    \includegraphics{../bld/figures/fig_counts_dp_jel_codes}
    \caption{Discussion Papers by Author \& Discussion Papers by JEL Codes}
    \label{fig:counts_dp_jel_codes}
\end{figure}
